### 2013-09-29 manuscripts/dpp-msbayes/dpp-msbayes.tex
More work on dpp ms.
Adding Intro, Discussion and notation table.

        \MakeUppercase{\large\bfseries Title}

@@

\section*{Introduction}
Being able to accurately infer the temporal distribution of divergences
across co-distributed taxa

Understanding evolutionary history is super cool.
Adding a comparative framework extra super cool.
Shared evolutionary history across taxa is often of interest:
processes of diversification and community assembly,
host-symbiont dynamics

@@

>>>>

        \MakeUppercase{\large\bfseries An Improved Approximate-Bayesian
            Model-choice Method for Estimating Shared Evolutionary History}

@@

\section*{Introduction}
Understanding the processes that generate biodiversity and regulate
community assembly is a major goal of Evolutionary Biology.
Given the dynamic nature of our planet, a major component of this goal is
understanding how geological and climatic processes can effect the
evolutionary history of whole groups of co-distributed species and their
associated microbiota, including pathogens.
Such large-scale processes predict patterns of shared evolutionary history,
e.g., temporally clustered divergences, across effected populations of
organisms.
At recent timescales, such events should leave a signature in the genetic
variation within and among the effected lineages.
Thus, being able to accurately estimate models of shared evolutionary
histories across co-distributed taxa would provide an important
tool for better understanding how regional and global biogeographic
processes affect biodiversity.

This inference problem is challenging due to the stochastic nature by which
mutations occur in populations and how they are inherited over generations
\citep{Hudson1990,WakeleyCoalescent}.
Thus, a method for estimating historical patterns of divergences across taxa
should explicitly model the stochastic mutational and ancestral processes that
generate and filter the genetic variation we observe in present-day genetic
data.
An appealing approach would be a comparative, Bayesian model-choice method for
inferring the probability of competing divergence histories while integrating
over uncertainty in mutational and ancestral processes via models of nucleotide
substitution and coalescence.

Such a method has been developed in an approximate-Bayesian model-choice
framework and is implemented in the software package \msb
\citep{Hickerson2006,Huang2011}.
Recently, however, \citet{Oaks2012} found the method can be strongly biased
toward spuriously estimating biogeographically interesting models of highly
clustered divergence times across taxa.
The results of their simulation-based assessment of the method suggest that a
combination of broad, uniformly distributed priors on nuisance parameters and a
U-shaped prior on divergence models were causing the bias.
The model implemented in \msb uses a discrete uniform prior over the number of
possible divergence events, which \citet{Oaks2012} show imposes a prior on
divergence models that puts most of the prior mass on models with either very
few or very many divergence times shared across the taxa.
Furthermore, the uniform priors on most of the model's nuisance parameters
likely hinder the marginal likelihoods of parameter-rich models, i.e., models
with many divergence time parameters \citep[this is often referred to as
Lindley's paradox;][]{Lindley1957}.
Thus, together, the priors on divergence models and nuisance parameters used in
\msb could create a strong bias toward models with very few divergence time
parameters.

Following the recommendations of \citet{Oaks2012} we modify the \msb model to
test whether alternative parameterizations and priors will improve the behavior
of the approximate Bayesian model-choice approach.
We implement a Dirichlet-process prior over all possible models of divergence,
and use gamma and beta probability distributions in place of uniform priors on
many of the model's parameters.
Our simulation-based analyses show that the new implementation has improved
robustness, accuracy, and power compared to the original model.
Also, our results show that both the uniform priors on nuisance parameters and
the U-shaped prior on divergence model contribute to the biases of \msb.

@@

%:TABLE-notation descriptions
    \scriptsize
    \addtolength{\tabcolsep}{-0.1cm}
\caption{Summary of the notation used throughout this work; modified from \citet{Oaks2012}.}
    \begin{tabular}{ l >{\raggedright\hangindent=0.5cm}m{14cm} }
        \textbf{Symbol} & \textbf{Description} \tn
        \npairs{} & Number of population pairs. \tn
        \popSampleSize{i}{} & The number of genome copies sampled from population pair $i$, with \popSampleSize{1}{i} sampled from population 1, and \popSampleSize{2}{i} from population 2. \tn
        \nloci{i} & Number of loci sampled from population pair $i$. \tn
        \nlociTotal & Total number of unique loci sampled. \tn
        \alignment{i}{j} & Sequence alignment of locus $j$ sampled from population pair $i$. \tn
        \alignmentSSObs{i}{j} & Population genetic summary statistics calculated from \alignment{i}{j}. \tn
        \alignmentVector & Vector containing the sequence alignments of each locus from each population pair: $(\alignment{1}{1},\ldots,\alignment{\npairs{}}{\nloci{\npairs{}}})$. \tn
        \ssVectorObs & Vector containing the summary statistics of each locus from each population pair: $(\alignmentSSObs{1}{1},\ldots,\alignmentSSObs{\npairs{}}{\nloci{\npairs{}}})$. \tn
        \ssSpace & Multi-dimensional Euclidean space around the observed summary statistics, \ssVectorObs. \tn
        \tol & Radius of \ssSpace, i.e., the tolerance of the ABC estimation. \tn
        \geneTree{i}{j} & Gene tree of the sequences in \alignment{i}{j}. \tn
        \geneTreeVector & Vector containing the gene trees of each locus from each population pair: $(\geneTree{1}{1},\ldots,\geneTree{\npairs{}}{\nloci{\npairs{}}})$. \tn
        \divTimeNum & Number of unique population divergence times among the \npairs{} population pairs. \tn
        \divTime{} & One of the \divTimeNum \iid divergence time paramters in \globalcoalunit generations. \tn
        \divTimeVector & Ordered set of divergence time parameters: $\{\divTime{1},\ldots,\divTime{\divTimeNum}\}$. \tn
        \divTimeIndex{i} & The index of the divergence time parameter in \divTimeVector that is mapped to population pair $i$. \tn
        \divTimeIndexVector & Vector of divergence time parameter indices: $(\divTimeIndex{1},\ldots,\divTimeIndex{\npairs{}})$. \tn
        \divTimeMap{i} & Time of divergence in \globalcoalunit generations between the populations of pair $i$. \tn
        \divTimeMapVector & Vector of divergence times for each of the population pairs: $(\divTimeMap{1},\ldots,\divTimeMap{\npairs{}})$. \tn
        \divTimeScaled{i}{j} & Scaled time of divergence between the populations of pair $i$ for locus $j$. \tn
        \divTimeScaledVector & Vector containing the scaled divergence times of each locus from each population pair: $(\divTimeScaled{1}{1},\ldots,\divTimeScaled{\npairs{}}{\nloci{\npairs{}}})$. \tn
        \descendantTheta{1}{i}, \descendantTheta{2}{i} & \myTheta{} of the $1^{st}$ and $2^{nd}$ descendent population, respectively, of pair $i$. \tn
        \ancestralTheta{i} & \myTheta{} of the population ancestral to pair $i$. \tn
        \descendantThetaVector{1}, \descendantThetaVector{2} & Vectors $(\descendantTheta{1}{1},\ldots,\descendantTheta{1}{\npairs{}})$ and $(\descendantTheta{2}{1},\ldots,\descendantTheta{2}{\npairs{}})$, respectively. \tn
        \ancestralThetaVector & Vector containing the \ancestralTheta{} parameters for each population pair: $(\ancestralTheta{1},\ldots,\ancestralTheta{\npairs{}})$. \tn
        \locusMutationRateScalar{j} & mutation-rate scaling parameter for locus $j$. \tn
        \locusMutationRateScalarVector & Vector containing the locus-specific mutation-rate parameters: $(\locusMutationRateScalar{1},\ldots,\locusMutationRateScalar{\nlociTotal})$. \tn
        \locusRateHetShapeParameter & Hyper-parameter for the shape of the gamma prior distribution on each \locusMutationRateScalar{}. \tn
        \bottleScalar{1}{i}, \bottleScalar{2}{i} & \myTheta{}-scaling parameters that determine the magnitude of the population bottleneck in the $1^{st}$ and $2^{nd}$ descendant population of pair $i$, respectively. The bottleneck in each descendant population begins immediately after divergence. \tn
        \bottleScalarVector{1}, \bottleScalarVector{2} & Vectors $(\bottleScalar{1}{1},\ldots,\bottleScalar{1}{\npairs{}})$ and $(\bottleScalar{2}{1},\ldots,\bottleScalar{2}{\npairs{}})$, respectively. \tn
        \bottleTime{i} & Proportion of time between present and \divt{i} when the bottleneck ends for the descendant populations of pair $i$. \tn
        \bottleTimeVector & Vector containing the \bottleTime{} parameters for each population pair: $(\bottleTime{1},\ldots,\bottleTime{\npairs{}})$. \tn
        \migrationRate{i} & Symmetric migration rate between the descendant populations of pair $i$. \tn
        \migrationRateVector & Vector containing the migration rates for each population pair: $(\migrationRate{1},\ldots,\migrationRate{\npairs{}})$. \tn
        \ploidyScalar{i}{j} & \myTheta{}-scaling constant provided by the user for locus $j$ of pair $i$. This constant is intended to allow the user to scale \myTheta{} for differences in ploidy among loci or differences in generation times among taxa. \tn
        \mutationRateScalarConstant{i}{j} & \myTheta{}-scaling constant provided by the user for locus $j$ of pair $i$. This constant is intended to allow the user to scale \myTheta{} for differences in mutation rates among loci or among taxa. \tn
        \ploidyScalarVector & Vector of ploidy and/or generation-time scaling constants: $(\ploidyScalar{1}{1},\ldots,\ploidyScalar{\npairs{}}{\nloci{\npairs{}}})$. \tn
        \mutationRateScalarConstantVector & Vector of mutation-rate scaling constants: $(\mutationRateScalarConstant{1}{1},\ldots,\mutationRateScalarConstant{\npairs{}}{\nloci{\npairs{}}})$. \tn
        \divTimeMean  & Mean of \divTimeMapVector. \tn
        \divTimeVar  & Variance of \divTimeMapVector. \tn
        \divTimeDispersion  & Dispersion index of \divTimeMapVector ($\divTimeVar/\divTimeMean$). \tn
        \numPriorSamples & Number of samples from the joint prior. \tn
        \paramSampleVector{} & Vector of parameter values drawn from the joint prior. \tn
        \ssVector{} & Vector containing the summary statistics calculated from data simulated under parameter values drawn from the prior (\hpvector{}). \tn
        \paramSampleMatrix{} & Random sample of $\paramSampleVector{1}, \ldots, \paramSampleVector{\numPriorSamples}$ drawn form the prior. \tn
        \ssMatrix & Summary statistic vectors $\ssVector{1}, \ldots, \ssVector{\numPriorSamples}$ for each $\hpvector{1}, \ldots, \hpvector{\numPriorSamples}$ drawn from the prior. \tn
    \label{tabNotation}
